Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-15 latex/paper.tex
added discussion of r2nmf scaling

Each iteration of \Cref{alg:parrank2nmf} incurs the same cost, so we analyze per-iteration computation and communication costs.
We first consider the cost of the Rank-2 NNLS solves, which are local computations.
In the notation of \Cref{alg:r2nnls}, matrix $\M{G}$ is $2\times 2$, so solving the unconstrained system (via Cholesky decomposition) and then choosing between single-positive-variables solutions if necessary requires constant time per row of $\M{C}$.
Thus, the cost of \cref{alg:r2nnls} is proportional to the number of rows of the first input matrix.
In the context of \Cref{alg:parrank2nmf}, the per-iteration computational cost of rank-2 solves is then $O((m+n)/p)$.
The other local computations are the matrix multiplications $\M[\hat]{W}^\Tra \M[\hat]{W}$ and $\M[\hat]{H}^\Tra \M[\hat]{H}$, which also amount to $O((m+n)/p)$ flops, and $\M[\hat]{A}^\Tra \M[\hat]{W}$ and $\M[\hat]{A} \M{H}$, which require $O(mn/p)$ flops because they involve the data matrix.
Thus, the computation cost is $\gamma \cdot O((mn+m+n)/p)$ and typically dominated by the multiplications involving $\M{A}$.
We track the lower order terms corresponding to NNLS solves because their hidden constants are larger than that of the dominating term.

@@

To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
Because the shape of the tree is data dependent and affects the overall costs, for the sake of analysis we will analyze only complete levels.
The number of rows in any node is $m$, the same as the root node, as each splitting corresponds to a partition of the columns.
Furthermore, because each split is a partition, every column of $\M{A}$ is represented exactly once in every complete level of the tree.
If we assume that all nodes perform the same number of NMF iterations ($\imath$) and power method iterations $(\jmath)$, then the dominating costs of a node with $\overline{n}$ columns is
\begin{equation*}
\label{eq:nodecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n} + \imath (m+\overline{n}) }{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) \\ + \alpha \cdot O((\imath+\jmath) \log p).
\end{split}
\end{equation*}
Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
\begin{equation*}
\label{eq:levelcost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn + \imath m 2^\ell}{p} \right) + \beta \cdot O((\imath+\jmath) n) \\ + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
\end{split}
\end{equation*}
Note that the only cost that depends on the level index $\ell$ is the latency cost.
Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
\begin{equation*}
\label{eq:treecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p}  \log k + \frac{\imath mk}{p} \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
\end{split}
\end{equation*}

@@

\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?  fixed number for each dataset?  any other experimental parameters we should specify?}

@@

@@

\subsection{Performance}

@@

\GB{4 plots: 2 data sets with speedup/breakdown}

\begin{itemize}
	\item theory says...
	\item Figs 3 and 5 show performance relative to the smallest number of nodes required to store data and factor matrices
	\item strong scaling is nearly linear, see 49X (check number) on 50 nodes for synthetic data and 7.3X (check number) on 80 nodes for image data
	\item Figs 4 and 6 show time breakdown, explain scaling: MM scales perfectly because it's mem-bw bound and the local problem size decreases linearly, bandwidth cost is fixed relative to $p$, so relative cost grows as we scale; comm is bandwidth bound because we see all-gather and reduce-scatter dominate all-reduce
\end{itemize}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0}
\label{fig:synrank2speedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0}
\label{fig:synrank2strongscaling}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
\caption{Strong Scaling Speedup for Rank-2 NMF \cref{alg:parrank2nmf} at Level-0 on Realworld Data}
\label{fig:rwrank2speedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
\caption{Relative Time Breakdown for Rank-2 NMF \cref{alg:parrank2nmf} Strong Scaling at Level-0 on Realworld Data}
\label{fig:rwrank2strongscaling}
\end{center}
\end{figure}

@@

\begin{itemize}
	\item theory says...t
	\item Fig 7 shows limit of strong scaling for synthetic data (time increases at 50 nodes compared to 40), Fig 8 gives breakdown and explains
	\begin{itemize}
		\item drastic performance dropoff may be artifact of all-gather, which should take about the same time as reduce-scatter, but speedup from 30 to 40 nodes is also slight
		\item by 50 nodes, about 50\% of time in communication; still mostly dominated by bandwidth cost because ag and rs dominate ar
		\item note the difference between the scaling of single rank-2 and entire tree: poor scaling must be due to lower levels, which we explore in the next section
	\end{itemize}
	\item Fig 9 shows reasonable scaling for entire tree -- speedup drops from 7x to 6x over 80 nodes; Fig 10 shows that MM is still 70\% of run time on 80 nodes, which allows for near-linear scaling
\end{itemize}

@@

\begin{itemize}
	\item theory says 
\end{itemize}

>>>>

Each iteration of \Cref{alg:parrank2nmf} incurs the same cost, so we analyze per-iteration computation and communication costs.
We first consider the cost of the Rank-2 NNLS solves, which are local computations.
In the notation of \Cref{alg:r2nnls}, matrix $\M{G}$ is $2\times 2$, so solving the unconstrained system (via Cholesky decomposition) and then choosing between single-positive-variables solutions if necessary requires constant time per row of $\M{C}$.
Thus, the cost of \Cref{alg:r2nnls} is proportional to the number of rows of the first input matrix.
In the context of \Cref{alg:parrank2nmf}, the per-iteration computational cost of rank-2 solves is then $O((m+n)/p)$.
The other local computations are the matrix multiplications $\M[\hat]{W}^\Tra \M[\hat]{W}$ and $\M[\hat]{H}^\Tra \M[\hat]{H}$, which also amount to $O((m+n)/p)$ flops, and $\M[\hat]{A}^\Tra \M[\hat]{W}$ and $\M[\hat]{A} \M{H}$, which require $O(mn/p)$ flops because they involve the data matrix.
Thus, the computation cost is $\gamma \cdot O((mn+m+n)/p)$ and typically dominated by the multiplications involving $\M{A}$.
We track the lower order terms corresponding to NNLS solves because their hidden constants are larger than that of the dominating term.

@@

To analyze the overall cost of the hierarchical clustering algorithm, we sum the costs over all nodes in the tree.
Because the shape of the tree is data dependent and affects the overall costs, for the sake of analysis we will analyze only complete levels.
The number of rows in any node is $m$, the same as the root node, as each splitting corresponds to a partition of the columns.
Furthermore, because each split is a partition, every column of $\M{A}$ is represented exactly once in every complete level of the tree.
If we assume that all nodes perform the same number of NMF iterations ($\imath$) and power method iterations $(\jmath)$, then the dominating costs of a node with $\overline{n}$ columns is
\begin{equation*}
\label{eq:nodecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) m\overline{n} + \imath (m+\overline{n}) }{p} \right) + \beta \cdot O((\imath+\jmath) \overline{n}) \\ + \alpha \cdot O((\imath+\jmath) \log p).
\end{split}
\end{equation*}
Because the sum of the number of columns across any level of the tree is $n$, the cost of the $\ell$th level of the tree is
\begin{equation}
\label{eq:levelcost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn + \imath m 2^\ell}{p} \right) + \beta \cdot O((\imath+\jmath) n) \\ + \alpha \cdot O((\imath+\jmath) 2^\ell \log p).
\end{split}
\end{equation}
Note that the only cost that depends on the level index $\ell$ is the latency cost.
Summing over levels and assuming the tree is nearly balanced and has height $O(\log k)$ where $k$ is the number of frontier nodes, we obtain an overall cost of \Cref{alg:hiernmf2} of
\begin{equation}
\label{eq:treecost}
\begin{split}
\gamma \cdot O\left( \frac{(\imath+\jmath) mn}{p}  \log k + \frac{\imath mk}{p} \right) + \beta \cdot O((\imath+\jmath) n \log k) \\ + \alpha \cdot O((\imath+\jmath) k \log p).
\end{split}
\end{equation}

@@

\GB{specify that $\imath=\jmath=100$ in all experiments, also what do we set $k$ to be?  fixed number for each dataset?  any other experimental parameters we should specify? could also specify these at the beginning of \Cref{sec:perf}}

@@

\GB{need shortcut names for datasets: synthetic, DC, and something better than IMAGE}

@@

\subsection{Performance}
\label{sec:perf}

@@

We perform strong scaling experiments for a single Rank-2 NMF (\Cref{alg:parrank2nmf}) on the synthetic and IMAGE datasets.
The theory (\Cref{eq:r2nmfcost}) suggests that perfect strong scaling is possible as long as the execution time is dominated by local computation.
Both the matrix multiplications and NNLS solves scale linearly with $1/p$ (we expect MatMul to dominate), but the bandwidth cost is independent of $p$ and the latency cost increases slightly with $p$.

\Cref{fig:synrank2speedup,fig:rwrank2speedup} show performance relative to the smallest number of nodes required to store data and factor matrices.
The synthetic data is chosen to fit on a single node, while the IMAGE data requires 10 nodes to store the input matrix as well as temporary and output matrices.
For these data sets, we observe nearly perfect strong scaling, with 47$\times$ speedup on 50 nodes (over 1 node) for synthetic data and 7.1$\times$ speedup on 80 nodes (over 10 nodes) for IMAGE data.

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_speedup.pdf}
\caption{Strong Scaling Speedup for Rank-2 NMF on synthetic data}
\label{fig:synrank2speedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/synthetic_rank2_strongscaling.pdf}
\caption{Relative Time Breakdown for Rank-2 NMF on synthetic data}
\label{fig:synrank2strongscaling}
\end{center}
\end{figure}

The relative time breakdowns are presented in \Cref{fig:synrank2strongscaling,fig:rwrank2strongscaling} and explain the strong scaling performance.
Each experiment is normalized to 100\% time, so comparisons cannot be readily made across numbers of compute nodes. 
For both data sets, we see that the time is dominated by MatMul, which is the primary reason for the scalability.
The dominant matrix multiplications are between a large matrix and a matrix with 2 columns, so it is locally memory bandwidth bound, with performance proportional to the size of the large matrix.
In each plot, we also see the relative time of all-gather and reduce-scatter increasing, which is because the local computation is decreasing while the communication cost is slightly increasing with $p$.
This pattern will continue as $p$ increases, which will eventually limit scalability, but for these data sets the MatMul takes around 80\% of the time at over 2000 cores.

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_speedup.pdf}
\caption{Strong Scaling Speedup for Rank-2 NMF on IMAGE data}
\label{fig:rwrank2speedup}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[height=2in, width=\columnwidth]{plots/realworld_rank2_strongscaling.pdf}
\caption{Relative Time Breakdown for Rank-2 NMF on IMAGE data}
\label{fig:rwrank2strongscaling}
\end{center}
\end{figure}

@@

\begin{itemize}
	\item theory (\Cref{eq:treecost}) says...
	\item Fig 7 shows limit of strong scaling for synthetic data (time increases at 50 nodes compared to 40), Fig 8 gives breakdown and explains
	\begin{itemize}
		\item drastic performance dropoff may be artifact of all-gather, which should take about the same time as reduce-scatter, but speedup from 30 to 40 nodes is also slight
		\item by 50 nodes, about 50\% of time in communication; still mostly dominated by bandwidth cost because ag and rs dominate ar
		\item note the difference between the scaling of single rank-2 and entire tree: poor scaling must be due to lower levels, which we explore in the next section
	\end{itemize}
	\item Fig 9 shows reasonable scaling for entire tree -- speedup drops from 7x to 6x over 80 nodes; Fig 10 shows that MM is still 70\% of run time on 80 nodes, which allows for near-linear scaling
\end{itemize}

@@

\begin{itemize}
	\item theory says per-level cost is \Cref{eq:levelcost}
\end{itemize}
